---
title: \vspace{-0.75cm} \begin{normalsize} Analysis of the Predictability of Weather Data \end{normalsize} \vspace{-0.5cm}
author:  \begin{normalsize} Ariane Stark \end{normalsize}
output: pdf_document
header-includes:
  \usepackage{setspace}
  \usepackage{float}
  \usepackage{array}
fontsize: 12pt
geometry: margin=1in
---
```{r include=FALSE}
source("R/Data-Analysis-Load.R")
```


\doublespacing

Temperature is determined by a myriad of factors. This paper seeks to analyze how useful historical weather data is in predicting. Specifically the analysis will focus on the weather in Amherst, Massachusetts from January 1, 2010 to December, 31 2021 for a total of `r nrow(data)` observations. The data has a seasonal trend (Figure 1a) with prior observations being heavily correlated with each other seen by looking at the sample autocorrelations (Figure 1b). 

```{r echo=FALSE, fig.height=3, fig.cap="Figure 1"}
cowplot::plot_grid(data.plot,acf.data.plot,nrow=1,labels = "auto",
                   rel_widths = c(4,3))
```


















































